---
title: Fitting the kinetic respiration model
#output: html_notebook
output: 
  rmarkdown::html_vignette: 
    keep_md: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Fitting the kinetic respiration model}
  %\usepackage[UTF-8]{inputenc}
---
```{r eval=FALSE, include=FALSE}
# twDev::genVigs()
rmarkdown::render("kinrespExperiment.Rmd","md_document") 
```

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra = 'style="display:block; margin: auto"'
    #, fig.align = "center"
    #, fig.width = 4.6, fig.height = 3.2
    , fig.width = 6, fig.height = 3.75 #goldener Schnitt 1.6
    , dev.args = list(pointsize = 10)
    , dev = c('png','pdf')
    )
knit_hooks$set(spar = function(before, options, envir) {
    if (before) {
        par( las = 1 )                   #also y axis labels horizontal
        par(mar = c(2.0,3.3,0,0) + 0.3 )  #margins
        par(tck = 0.02 )                          #axe-tick length inside plots             
        par(mgp = c(1.1,0.2,0) )  #positioning of axis title, axis labels, axis
     }
})
# if (!require(ggplot2) || !require(dplyr) || !require(purrr)) {
# 	print("To generate this vignette, ggplo2, dplyr, and purrr are required.")
# 	exit(0)
# }
#themeTw <- theme_bw(base_size = 10) + 
#  theme(axis.title = element_text(size = 9))
```

## Experimental data

Here we take respiration time series from the dataset in Wutzler 2012 of one
experiment consisting of incubation of three replicate soil samples.

```{r, message = FALSE}
library(twKinresp, quietly = TRUE) 
rde <- subset(respWutzler10, suite == "Face" & experiment == 9 )
summary(rde)
```

```{r}
plot( resp ~ time, data = rde, col = rde$replicate
      , xlab = "Time (hour)", ylab = "Respiration (gC/gCsoil/hour)")
```

First, we constrain each time series to the unlimited exponential growth phase.
The package achieves this byfitting a three parameter exponential equation 
to successively
shortened time series. A deviation from the exponential model introduces 
positive correlations in the model-data residuals.
Hence, the longest time series is selected
where there was no positive correlation in the model-data residuals 
between the mimimum and the last last point.

This works quite well with data that displays a typical exponential growth phase
but fails if there are deviations from exponential growth present, such as 
small oscillations in the respiration. Then the records have to be 
selected manually.

```{r, message = FALSE}
res4 <- kinrespGrowthphaseExperiment(rde, weights = varPower(fixed = 0.5) )
rde.e <- getUnlimitedGrowthData(res4)
plot( resp ~ time, data = rde.e, col = rde.e$replicate
      , xlab = "Time (hour)", ylab = "Respiration (gC/gCsoil/hour)")
```

## Mixed linear models

We are interested of the microbial growth parameters for the soil, i.e. 
across the replicates.

There are several suboptimal approaches to aggregate across replicates.

- Fitting a single growth curve to all the data points
- Computing the mean respiration for each time and 
  fitting a model to the single aggregated series
- Fitting a growth curve to each replicate, and then taking the average across
  the parameters
  
The last approach is wrong because the growth curve of the averaged parameters
is not an average growth curve.

The two first approaches will provide a good average growth curve but provide
wrong uncertainty bounds. The first approach neglects the groupings, i.e. 
dependency between the observations of one replicate 
and underestimates uncertainty.
The second approach drops the advantage of having multiple replicates and
can overestimate uncertainty.

Nonlinear mixed effects models provide an proper statistical approach that 
uses all of the records and accounts for the groupings ob observations 
of a replicate.
This approach is implemented in this package.

## Fitting and inspecting the mixed effects kinetic respiration model

Several versions with random effects in different parameters are fitted,
and the model with lowest AIC is selected.
```{r}
res5Scen <- fitKinrespExperiment(
  #,showFitErrorMsg = TRUE
  rde.e, coefKinresp(res4,rde.e), weights = varPower(fixed = 0.5))
res5Scen$aics
```

In the example, the variant with the lowest AIC had a random effect
in initial biomass, $x_0$, and initial proportion of active biomass, $r_0$,
indicating that those properties differed across replicates.

The variant with a random effect in additionally the maximum growth rate, 
$\mu_{max}$, could not be fitted to the data (AIC==NA). This indicates either 
no differences between the replicates or not enough replicates to resolve 
their differences.

A plot of the fits to the individual replicates shows a good agreement
between the model and the observations.
```{r fits, echo=FALSE}
  rde.e$fitted <- fitted(res5Scen$model)
  plot( resp ~ time, data = rde.e, col = rde.e$replicate 
        , xlab = "Time (hour)", ylab = "Respiration (gC/gCsoil/hour)")
  tmp <- by( rde.e, rde.e$replicate, function(rder){
    lines(fitted~time,data = rder, col = as.numeric(rder[1,replicate]))})
```

The selected model object is stored in entry `model` of the result.

```{r}
kFit = res5Scen$model
fixef(kFit)
```
## Transformed and original scale

The model parameters $x_0$ and $\mu_{max}$ are constrained to be strictly 
positive and their confidence interval should not include values below zero.
Therefore, log-transformed versions, $x_{0l}$ and $\mu_{maxl}$ are fitted 
with the model. This corresponds to assuming that they are log-normally 
distributed. The values and distribution quantiles at original scale are simply
found by taking the exponential of their values.

Similarly, model parameter $r_0$ is constrained in interval $(0,1)$. Hence,
the model fits a logit-transformed version $r_{0l}$, and the values at original
scale are found by applying the `invlogit` function.

To simplify the process, function `coefKinresp` and `confintKinresp` 
apply these conversions.

```{r}
#exp(fixef(kFit)["x0l"])
#invlogit(fixef(kFit)["r0l"])
coefKinresp(fixef(kFit))
```

The uncertainty bounds of the population grwoth parameters at original scale are:
```{r}
confintKinresp(confint(kFit))
```

The help page `?coefKinresp` links to further functions accessing
the fitted results at transformed and original scale.

Note, that the expected value of a lognormal distribution depends on
both, the location, $\mu$, and the shape parameter, $\sigma$, and is larger than
the exponential of $\mu$:

$$
E(x) = e^{\mu + \sigma^2/2}
$$

Function `kinrespParDist` computes this expectec value and other statistics.
We recommend reporting the distribution parameters $(\mu,\sigma)$ along 
with the expected value and the confidence bounds.
```{r}
pars <- kinrespParDist(kFit)
pars[,c("mean","cf025","cf975","mu","sigma")]
```

The skewness is vizalized by the following plot of the density distribution 
of initially active fraction $r_0$.

```{r r0Density, echo=FALSE}
iPar = "r0"  
xGrid <- seq( pars[iPar,"cf025"]*0.8, pars[iPar,"cf975"]*1.2, length.out = 80)
#fx <- dlnorm(xGrid, mean = pars[iPar,"mu"],sd = pars[iPar,"sigma"])
fx <- dlogitnorm(xGrid, mu = pars[iPar,"mu"],sigma = pars[iPar,"sigma"])
plot( fx ~ xGrid, type = "l", xlab = iPar, ylab = "density" )
cols = c("red","green","blue","gray","gray")
abline(v = pars[iPar,c("mle","median","mean","cf025","cf975")]
       , col = cols)
legend("topright", c("density","mode","median","mean","95% bounds")
       ,col = c("black",cols), lty = "solid")
```

